Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2021-03-20 manuscript/v2_medrxiv/manuscript.tex
fix manuscript v2?

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing

@@

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}. It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf. \citep{gencoglu2020causal}). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks). Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}. For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets. A further analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}. We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).

@@

Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

@@

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}. It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks). Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}. For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets. A further analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}. We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).

@@

Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
